---
title: Outbreak of gastroenteritis after a high school dinner in Copenhagen, Denmark,November
  2006
author: "Zhian N. Kamvar, Janetta Skarp, Alexander Spina, and Patrick Keating"
output:
  html_document
---

The following code has been adapted to *R* for learning purposes. The initial contributors and copyright license are listed below. All copyrights and licenses of the original document apply here as well. 

**Contributors to *R* code:**  
Daniel Gardiner(PHE) and Lukas Richter (AGES)

# Copyright and license
This case study was designed under an ECDC service contract for the development of training material (2010). The data were slightly modified for training purposes.

**Source :**
This case study is based on an investigation conducted by Jurgita Pakalniskiene, Gerhard Falkenhorst (Statens Serum Institut, Copenhagen) and colleagues

**Authors:**
Jurgita Pakalniskiene, Gerhard Falkenhorst, Esther Kissling, Gilles Desvé.

**Reviewers:**
Marta Valenciano, Alain Moren.

**Adaptions for previous modules:**
Irina Czogiel, Kristin Tolksdorf, Michaela Diercke, Sybille Somogyi, Christian Winter, Sandra Dudareva-Vizule with the help of Katharina Alpers, Alicia Barrasa, Androulla Efstratiou, Steen Ethelberg, Annette Heißenhuber, Aftab Jasir, Ioannis Karagiannis and Pawel Stefanoff


**You are free:**

+ to Share - to copy, distribute and transmit the work
+ to Remix - to adapt the work
Under the following conditions:
+ Attribution - You must attribute the work in the manner specified by the author or licensor (but not in any way that suggests that they endorse you or your use of the work). The best way to do this is to keep as it is the list of contributors: sources, authors and reviewers.
+ Share Alike - If you alter, transform, or build upon this work, you may distribute the resulting work only under the same or similar license to this one. Your changes must be documented. Under that condition, you are allowed to add your name to the list of contributors.
+ You cannot sell this work alone but you can use it as part of a teaching.
With the understanding that:
+ Waiver - Any of the above conditions can be waived if you get permission from the copyright holder.
+ Public Domain - Where the work or any of its elements is in the public domain under applicable law, that status is in no way affected by the license.
+ Other Rights - In no way are any of the following rights affected by the license:
+ Your fair dealing or fair use rights, or other applicable copyright exceptions and limitations;
+ The author's moral rights;
+ Rights other persons may have either in the work itself or in how the work is used, such as publicity or privacy rights.
+ Notice - For any reuse or distribution, you must make clear to others the license terms of this work by keeping together this work and the current license.
This licence is based on http://creativecommons.org/licenses/by-sa/3.0/

```{r setup, include = FALSE}
# This tells knitr to start everything from the project directory
knitr::opts_knit$set(root.dir = here::here())
```


# Descriptive analysis in *R*

## Preparing packages and data

```{r load_packages, echo = TRUE, results='hide', message=FALSE, warning=FALSE}
library("ggplot2")
library("skimr")
library("Hmisc")
library("epiR")
library("incidence")
library("here")
```


```{r read_clean_data}
# read in your data from a csv file 
# Select separator as comma (sep=",")
# do not import 'string' variables as 'Factors' (stringsAsFactors=FALSE) 
# Factors are a special datatype, covered later - character variables are simpler
# data frame read in and saved in R as "cph"


cph <- read.csv(here::here("data", "copenhagen_descriptive.csv"),
                stringsAsFactors = FALSE) 
```




### Dataset description and tabulation   

### Describe signs and symptoms of cases 

We will use 2x2 tables with percentages to describe the signs and symptoms of 
all the cases. To do this, we must tabulate the counts, calculate the proportions
of those counts, and then combine those into a single table. Here's an example
of the table for cases that have diarrhoea:

```{r describe_signs_and_symptoms}
# create a frequency table diarrhoea among cases
a <- table(cph$diarrhoea[cph$case == 1])

# frequencies divided by total cases (as proportion) and rounded to one decimal place
props <- round(prop.table(a) * 100, digits = 1)

# bind a column (cbind) to the above table with:
b <- cbind(n = a, prop = props)
b
```

However, we have a lot of variables we want to do this over, it's better to
write a function that can do this for us over and over again. The key to writing
a function to to look at a procedure and think about what parts stay the same
and what parts change. In the case of our 2x2 table, the only thing that changes
is the vector we want to describe. If we replace that by a variable called
"x", we can write our function like so:

```{r describe_signs_and_symptoms_fun}
twobytwo <- function(x) {
  # create a frequency table diarrhoea among cases
  a <- table(x)

  # frequencies divided by total cases (as proportion) and rounded to one decimal place
  props <- round(prop.table(a) * 100, digits = 1)

  # bind a column (cbind) to the above table with:
  b <- cbind(n = a, prop = props)
  b
}
```

> **n.b.** This function is *portable*, meaning that I can copy it to any other
> R script and it will work exactly the same. 

Now that we have our function, we can create the table with just one line:

```{r use_fun}
twobytwo(cph$diarrhoea[cph$case == 1])
twobytwo(cph$bloody[cph$case == 1])
```

Of course, we have several variables, we don't necessarily want to write this
over and over, so we will use a function called `lapply()`. This function takes
a data frame, list, or vector, applies a function over each variable separately,
and then returns a list of results. If we were to use `lapply()` to get the 
results from above, we might do this:

```{r use_lapply_1}
vars <- c("diarrhoea", "bloody")
lapply(cph[cph$case == 1, vars, drop = FALSE], FUN = twobytwo)
```

Now you try it with the following variables:

```{r lapply_vars}
vars <- c("diarrhoea", "bloody", "vomiting", "abdo", "nausea", "fever", "headache", "jointpain")
```




```{r describe_part_2, echo = FALSE}
  # Version 2

# define the variables which you would like to look at and save as "vars"
vars <- c("diarrhoea", "bloody", "vomiting", "abdo", "nausea", "fever", "headache", "jointpain")
lapply(cph[cph$case == 1, vars, drop = FALSE], twobytwo)
```


### Determine the median incubation period 

```{r median_incubation}
summary(cph$incubation[cph$case == 1])
```

### Describe the cohort in terms of person 

```{r person_cohort}
# look at age and other variables of interest
Hmisc::describe(cph$age)
```


### Calculate the overall attack rates as well as the attack rates stratified by person characteristic  

Remember, you can also refer to a dataset using square brackets, the part before a comma refers to the rows and after refers to columns (named or numerically). Or visually: `cph[row, column]`.
For example: `cph[ , "age"]` gives you the variable age as a vector, or `cph[2:4 , ]` gives you rows two to four. 


This is a very similar procedure to the 2x2 tables above, except now we want to include the non-cases. 
We'll demonstrate it for "group" first:

```{r attack-rate-group}
tab  <- table(cph$group, cph$case)

# create a table of proportions
prop <- round(prop.table(tab, margin = 1) * 100, digits = 1)

# bind the column with AR from b to the 2by2 table in a
res  <- cbind(tab, prop[, 2, drop = TRUE])

# rename your columns
colnames(res) <- c("non case", "case", "AR%")
res
```

Much like the function we used above, only one thing changes: the variable of
interest, but we also need to have a vector of cases to complete the table.
Since functions should be stand-alone, we will create another variable called
"case":

```{r attack-rate-fun}
AR <- function(x, case) {
  tab  <- table(x, case)

  # create a table of proportions
  prop <- round(prop.table(tab, margin = 1) * 100, digits = 1)

  # bind the column with AR from b to the 2by2 table in a
  res  <- cbind(tab, prop[, 2, drop = TRUE])

  # rename your columns
  colnames(res) <- c("non case", "case", "AR%")
  res
}
```

We can see that it works for group:

```{r attack-rate-group-2}
AR(cph$group, case = cph$case)
```


Now we can use it with lapply, supplying our "case" argument after the function
name. This is explained in the help documentation for "lapply"

```{r attack-rate-lapply}
# students and sex
vars <- c("group", "class", "sex")
lapply(cph[vars], AR, case = cph$case)
```

To see how to combine this code to create a ready-to-present table for exporting see the appendix. 


# Describing time in *R*


### Recoding for date compatability 

Date compatibility is always an issue when switching between softwares, R sometimes has a problem reading in dates from excel files, however there is user-written code which fixes these kinds of glitches. 

In the current dataset it is simply a case of creating useful date variables by some data manipulation. 

The variable *start* is an example of bad data collection. 
Nevertheless it consists of numbers (1, 2 and 3); these numbers correspond to dates (11th, 12th and 13th of November 2006).

```{r}
# Create a useful date variable from start
# set yourself a reference date (one day before onset)
refdate <- as.Date("2006-11-10")

# use start to create dates from your start variable
cph$dayonset <- as.Date(refdate + cph$start)
```


### Construct epicurve by day
It is possible to create an epicurve manually using base-R code (see appendix), however for simplicity, there is a user written function (Daniel Gardener, PHE). 
The epicurve function allows creation of easily formatted epicurves. To find out more about the function, first load it as above and then click on function in the Global Environment tab on the right of the R Studio window.
This function uses the ggplot package (use ??ggplot2 to find out more about this function) and so is quite easily manipulated and saved after being created.



```{r epicurve, warning = FALSE}
cph_incidence <- incidence(cph$dayonset, interval = 1)
plot(cph_incidence)
```




