---
title: Outbreak of gastroenteritis after a high school dinner in Copenhagen, Denmark,November
  2006
author: "Zhian N. Kamvar, Janetta Skarp, Alexander Spina, and Patrick Keating"
output:
  html_document
---

The following code has been adapted to *R* for learning purposes. The initial contributors and copyright license are listed below. All copyrights and licenses of the original document apply here as well. 

**Contributors to *R* code:**  
Daniel Gardiner(PHE) and Lukas Richter (AGES)

# Copyright and license
This case study was designed under an ECDC service contract for the development of training material (2010). The data were slightly modified for training purposes.

**Source :**
This case study is based on an investigation conducted by Jurgita Pakalniskiene, Gerhard Falkenhorst (Statens Serum Institut, Copenhagen) and colleagues

**Authors:**
Jurgita Pakalniskiene, Gerhard Falkenhorst, Esther Kissling, Gilles Desvé.

**Reviewers:**
Marta Valenciano, Alain Moren.

**Adaptions for previous modules:**
Irina Czogiel, Kristin Tolksdorf, Michaela Diercke, Sybille Somogyi, Christian Winter, Sandra Dudareva-Vizule with the help of Katharina Alpers, Alicia Barrasa, Androulla Efstratiou, Steen Ethelberg, Annette Heißenhuber, Aftab Jasir, Ioannis Karagiannis and Pawel Stefanoff


**You are free:**

+ to Share - to copy, distribute and transmit the work
+ to Remix - to adapt the work
Under the following conditions:
+ Attribution - You must attribute the work in the manner specified by the author or licensor (but not in any way that suggests that they endorse you or your use of the work). The best way to do this is to keep as it is the list of contributors: sources, authors and reviewers.
+ Share Alike - If you alter, transform, or build upon this work, you may distribute the resulting work only under the same or similar license to this one. Your changes must be documented. Under that condition, you are allowed to add your name to the list of contributors.
+ You cannot sell this work alone but you can use it as part of a teaching.
With the understanding that:
+ Waiver - Any of the above conditions can be waived if you get permission from the copyright holder.
+ Public Domain - Where the work or any of its elements is in the public domain under applicable law, that status is in no way affected by the license.
+ Other Rights - In no way are any of the following rights affected by the license:
+ Your fair dealing or fair use rights, or other applicable copyright exceptions and limitations;
+ The author's moral rights;
+ Rights other persons may have either in the work itself or in how the work is used, such as publicity or privacy rights.
+ Notice - For any reuse or distribution, you must make clear to others the license terms of this work by keeping together this work and the current license.
This licence is based on http://creativecommons.org/licenses/by-sa/3.0/


# Univariable analysis (Cohort and case control)

To identify (the) potential vehicle(s) in the outbreak, proceed with an analytical study where you use statistical tests to investigate the associations of some suspicious food items with the disease.


If you haven't completed the previous section then you can load the following dataset for the next steps.


```{r read_univariable_data}
# read in your data from a csv file 
# Select separator as comma (sep=",")
# do not import 'string' variables as 'Factors' (stringsAsFactors=FALSE) 
# Factors are a special datatype, covered later - character variables are simpler
# data frame read in and saved in R as "cph"


cph <- read.csv(here::here("data", "copenhagen_univariable.csv"), stringsAsFactors = FALSE) 
```



There is more than one way to create univariable tables.  One gives you a two by two and quite a lot of info, and the other is user-written code which will provide you with nice output like STATA's cstable. 



Think about which variables you might want to include when checking for effect modification or confounding. One common strategy is to base this decision on the univariable results obtained and a p-value threshold of 0.25. Also, food items that are known risk factors for gastroenteritis could also be included regardless of their univariable p-value.


## Detailed output (method 1)

In order to use the `epi.2by2()` function, we first need to convert the outcome and exposure variables into factor variables to facilitate interpretation.  

Outcome and exposure variables of interest need to be factor variables prior to using the function, in order to be relevelled from (0, 1) to (1, 0) so that they can be correctly organised in 2-by-2 tables.

```{r convert_factors}
# We list the outcome/exposure variables
vars <- c("shrimps", "veal", "pasta", "sauce", "champagne", "rocket", "case")



# Convert all of those variables to factor variables and re-order the levels to aid interpretation
for (var in vars) {
  cph[[var]] <- factor(cph[[var]], levels = c(1, 0))
}
```


The `epi.2by2()` function from the *epiR* package can be used to calculate both RRs and ORs.
You can find out more information on the function by writing **?epi.2by2** in the console. 
The epi.2by2 function requires data to be in a table format. 
We can specify that we want to calculate RRs or ORs by adding **method = "cohort.count"** or **method = "case.control"**, respectively. 
You can do that first for looking at the risk for being a case by exposure to shrimp in a cohort study design.  




```{r get_2by2}
# Create a table with exposure and outcome variables
counts <- table(cph$shrimps, cph$case)

# Apply epi.2by2 function to the table
shrimp <- epiR::epi.2by2(counts, method = "cohort.count")

# to view the output
shrimp
```


This example involves using the `epi.2by2()` function from the *epiR* package. To make things faster (but not necessarily easier to understand) it has been run in a for-loop and the output has been hidden because it is very long. 

With extra code you could extract information from epi.2by and piece together an output table like in STATA.
Here we have just chosen to show this in the console (extraction will be shown later in the case study).

```{r lotta_2by2s, results="hide"}

## vector of variables you want to run in the for-loop
vars <- c("pasta", "veal", "champagne", "sauce", "shrimps")

# for each variable run a table and then get the rr out using epi.2by2 from the epiR package

for (var in vars) {
  # create table with the variable of interest by case classification
  a <- table(cph[, var], cph$case)
  
  # pass your counts table to epi.2by2
  test <- epi.2by2(a,  method = "cohort.count")
  
  # print the output of the table in the console and label which variable it is
  print(var)
  print(test)
}
```







## Simplified output (method 2)

A much more straight forward way is to simply apply a user-written function which requires much less input from you. 
The *sva* function (Daniel Gardner, PHE) basically integrates the steps from above to create a nice output table. 

```{r run_sva_fun}
# You already sourced the function at the begining of this case study 
# if you havent then type: source(here("scripts/single.variable.analysis.v0.2.R"))
# nb. click on "sva" in your global environment to view source code and read explanations

# cohort study
sva(cph, outcome = "case", exposures = vars, measure = "rr", verbose = TRUE)

# cohort count
sva(cph, outcome = "case", exposures = vars, measure = "or", verbose = TRUE)
```


# Different statistical tests in *R*


Note that for the below tests, no tables are printed alongside, however you could create these tables using the table and propstable (for percentages) functions. 


## Is gender associated with being a case? 

```{r tests}
# using a chi-squared test
  # chisq.test funciton requires you to input a table
chisq.test(table(cph$sex, cph$case))


# using a fisher's exact test
fisher.test(table(cph$sex, cph$case))
```



## Is class associated with being a case?
Here you can either compare proportions, using the chi-squared test, or use the Wilcoxon-Mann-Whitney test to compare distributions. 
For the Wilcoxon-Mann-whitney test to work, all variables need to be numeric - however it is not possible to go directly from a factor to a numeric. The intermediate is to turn it in to a character and then to a numeric. 
```{r tests2}

# Using chi-squared
chisq.test(table(cph$class, cph$case))

# using the wilcoxon test 
  # this is a ranksum test and the function pulls the counts numbers by itself

# change case from a factor to numeric
cph$case <- as.numeric(as.character(cph$case))

# run the wilcoxon 
wilcox.test(cph$class, cph$case)

```

## Is there a difference between age in males and females? 
```{r test_sex}


# Using a wilcox test
wilcox.test(cph$age, cph$sex)

# Using a t-test to see difference in mean if normally distributed 
  # shapiro tests if significantly different from normal distribution
shapiro.test(cph$age)

t.test(cph$age ~ cph$sex)


```




## Is there a dose-response relationship between the food items and being sick? 

Look at the food items that seem most suspicious to you. 
Hint: it's pasta and veal

There is two possible ways to do this. The first is to use the wilcox.test again, and the second is to calculate ORs using the `epi.2by2()` function. Using the `epi.2by2()` function for dose responses is a bit more dense (traditionally this would be done using regression), so we have not shown it here; if you would like to see the code for this it is available in the appendix. 



```{r test_the_veal}


vars <- c("vealD", "pastaD")



for (var in vars) {
  # selects only those cases where dose response is filled out (not empty) and saves as a dataset
  dataset <- cph[is.na(cph[var]) == FALSE, ]
  
  # saves in "form" as a text
  # e.g. "pastaD ~ case" which can then be inserted in the wilcox.test to get dose response
  form <- as.formula(paste0(var, "~", "case"))
  
  # put your formula above in to wilcox using the reduced dataset
  a <- wilcox.test(form, data = dataset)
  
  # show a table with doses and case and the wilcox output
  print(var)
  print(table(cph[, var], cph$case))
  print(a)
}

```


